Comparison Of Several Sequential And Parallel Derivative-free Global Optimization Algorithms

### 2019-03-31 2019/PAVT/slides/slides.tex
PAVT: the final version of the slides

\begin{frame}
  \frametitle{Methods}
  \begin{itemize}
    \item[$\square$] \textbf{Deteriminstic}
    \begin{itemize}
      \item Have complicated internal structure in multidimensional case;
      \item Usually store and use the whole history of trials accumulated during search;
      \item Require non-trivial parallelization schemes;
      \item Under some assumptions convergence to the global solution is guaranteed;
    \end{itemize}
    \item[$\square$] \textbf{Stochastic}
    \begin{itemize}
      \item Have relative simple internal structure;
      \item Require constant amount of memory to store internal state of some random process or individuals of population;
      \item In most cases allows wide-scale parallelization by trials or by running many copies of the same method
      with different seeds in parallel.
      \item Convergence is guaranteed in probabilistic sense only.
    \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Methods}
  In this work considered the following solvers available in open-source:
  \begin{itemize}
    \item[$\square$] \textbf{Deteriminstic}
    \begin{itemize}
      \item DIRECT, DIRECT$l$;
      \item AGS, AGS$l$;
    \end{itemize}
    \item[$\square$] \textbf{Stochastic}
    \begin{itemize}
      \item Multi Level Single Linkage;
      \item StoGO;
      \item Differential Evolution;
      \item Controlled Random Search;
      \item Dual Simulated Annealing;
      \item Ant colony optimization;
    \end{itemize}
  \end{itemize}
\end{frame}

@@

\begin{frame}
  \frametitle{Basic optimization method}
  Optimization method generates search sequence \(\{x_k\}\) and consists of the following steps:
  \begin{enumerate}
    \setlength{\itemindent}{.1in}
    \item[Step 1.] Sort the search information (one-dimensional points) in increasing order.
    \item[Step 2.] For each interval \((x_{i-1}, x_i)\) compute quantity \(R(i)\), called characteristic.
    \item[Step 3.] Choose \(p\) intervals \((x_{t_j-1}, x_{t_j})\) with the greatest characteristics and
    compute objective \(\varphi(y(x^{k+j}))\) in points chosen using the decision rule \(d\):
    \begin{displaymath}
      x^{k+1+j}=d(t)\in (x_{t_j-1}, x_{t_j}),\:j=\overline{1,p}
    \end{displaymath}
    \item[Step 4.] If \(x_{t_j}-x_{t_j-1}<\varepsilon\) for one of \(j=\overline{1,p}\), stop the method.
  \end{enumerate}
  \textit{\footnotesize	{Detailed description: Strongin R.G., Sergeyev Ya.D.: Global optimization with non-convex constraints. Sequential and parallel algorithms (2000), Chapter 7}}
\end{frame}

@@

\begin{frame}
  \frametitle{Parallel Globalizer solver}
  \begin{figure}[ht]
    \includegraphics[width=0.95\textwidth]{evolvents_parallel.pdf}
  \end{figure}
\end{frame}

\begin{frame}
  \frametitle{Parallel MIDACO solver}
\end{frame}

@@

\begin{frame}
  \frametitle{Results of applying the parallel algorithm}
  \begin{table}
    \centering
    \caption{Averaged numbers of iterations executed by the parallel algorithms for solving the test
  optimization problems}
    \label{tab:iterations}
    \begin{tabular}{cccccccc}
      \cline{3-8}\noalign{\smallskip}
      \multicolumn{2}{c}{  } & \textit{p} & \multicolumn{2}{c}{Globalizer} & &
  \multicolumn{2}{c}{MIDACO}   \\
      \noalign{\smallskip} \cline{4-5} \cline{7-8}  \noalign{\smallskip}
      \multicolumn{2}{c}{  } & & \textit{Simple} & \textit{Hard} & & \textit{Simple} &
  \textit{Hard}  \\
      \noalign{\smallskip}\hline
      I & \textbf{1 cluster node}
        & \textit{1} &   25270 (100)  & 55180 (99) & & 27645 (98) & 72068 (71)  \\
      &  & \textit{16} & 1765 (100)  & 3714 (100) & &  1640 (97) & 4304 (70) \\
      \hline \noalign{\smallskip}
  II  & \textbf{2 cluster nodes}  %\multirow{3}{*}{}
    & \textit{1} & 13056 (100) & 22938 (99) & & 10558 (89) & 27128 (73) \\
  &   & \textit{16} & 732 (100) & 1759 (100)  &  & 1130 (92) & 2254 (73) \\
      \hline \noalign{\smallskip}
  III & \textbf{4 cluster nodes} %\multirow{3}{*}{}
    & \textit{1}  & 5016 (100) & 12703 (100) & & 5777 (94) & 15980 (75) \\
  & & \textit{16} & 367 (100) & 776 (100) & & 529 (88) &  1264 (66) \\
      \hline \noalign{\smallskip}
  VI & \textbf{8 cluster nodes} %\multirow{3}{*}{}
    & \textit{1}  & 2103 (100) & 5063 (100) & & 2847 (97) & 9853 (83)\\
  & & \textit{16} & 145 (100)  & 310 (100)  & & 272 (83) & 774 (57)\\
      \hline \noalign{\smallskip}
  V & \textbf{12 cluster nodes} %\multirow{3}{*}{}
    & \textit{1}  & 1155 (100) & 2399 (100) & & 2233 (98) & 6022 (86) \\
  & & \textit{16} & 76 (100)  & 159 (100)  & & 168 (87) & 393 (53)\\
      \noalign{\smallskip}\hline
    \end{tabular}
  \end{table}
\end{frame}

\begin{frame}
  \frametitle{Results of applying the parallel algorithm}
  \begin{table}
    \centering
    \caption{Speedup of parallel computations executed by the parallel algorithms}
    \label{tab:speedup}
    \begin{tabular}{cccccccc}
      \cline{3-8}\noalign{\smallskip}
      \multicolumn{2}{c}{  } & \textit{p} & \multicolumn{2}{c}{Globalizer} & &
  \multicolumn{2}{c}{MIDACO}   \\
      \noalign{\smallskip} \cline{4-5} \cline{7-8}  \noalign{\smallskip}
      \multicolumn{2}{c}{  } & & \textit{Simple} & \textit{Hard} & & \textit{Simple} &
  \textit{Hard}  \\
      \noalign{\smallskip}\hline
  I  & \textbf{1 cluster node}  %\multirow{3}{*}{}
      & \textit{1}   & 25270 (27.3s) & 55180 (61.8s) & & 27645 (32.2s) & 72068 (132.5s)  \\
    &  & \textit{16} & 14.3(11.7) & 14.9(12.6)  & &  16.9(14.4) & 16.7(14.4) \\
    \hline \noalign{\smallskip}
  II  & \textbf{2 cluster nodes}  %\multirow{3}{*}{}
    & \textit{1}      &   1.9(1.9) & 2.4(2.2)  & & 2.6(1.7) & 2.7(2.3) \\
    &   & \textit{16} & 34.5(20.4) & 31.4(18.8) & & 24.5(18.8) & 32.0(29.1) \\
      \noalign{\smallskip}\hline	\noalign{\smallskip}
  III  & \textbf{4 cluster nodes}  %\multirow{3}{*}{}
  & \textit{1}      & 5.0(4.6) & 4.3(4.1) & & 4.8(3.9) & 4.5(4.4) \\
  &   & \textit{16} & 68.8(23.7) & 71.2(25.8) & & 52.3(32.9) & 57.0(50.2) \\
    \noalign{\smallskip}\hline	\noalign{\smallskip}
  VI & \textbf{8 cluster nodes} %\multirow{3}{*}{}
  & \textit{1}    & 12.0(8.7) & 10.9(8.3) & & 9.7(8.2)    & 7.3(9.0)    \\
  & & \textit{16} & 174.1(4.7) & 177.8(5.3) & & 101.6(51.6) & 93.1(84.1) \\
      \noalign{\smallskip}\hline
      V & \textbf{12 cluster nodes} %\multirow{3}{*}{}
      & \textit{1}    & 21.9(11.4)  & 23.0(12.8)  & & 12.4(11.1)  & 12.0(14.8)  \\
      & & \textit{16} & 333.5(2.7) & 347.9(3.0) & & 164.5(82.0) & 183.2(129.7) \\
          \noalign{\smallskip}\hline
    \end{tabular}
  \end{table}
\end{frame}

\begin{frame}
  \frametitle{Results of applying the parallel algorithm}
  \begin{figure}[ht]
    \centering
    \subfloat[$p=1$]{{\includegraphics[width=.5\textwidth]{speedup_midaco_globalizer_p_1.pdf}}\label{fig:speedup_p1}}
    \subfloat[$p=16$]{{\includegraphics[width=.5\textwidth]{speedup_midaco_globalizer_p_16.pdf}}\label{fig:speedup_p16}}
    \caption{Speedup demonstrated by the parallel algorithms when solving problems from the GKLS 4d Hard class}
  \end{figure}
\end{frame}

\begin{frame}
  \frametitle{Results of applying the parallel algorithm}
  \begin{figure}[ht]
    \centering
    \subfloat[$p=1$]{{\includegraphics[width=.5\textwidth]{time_midaco_globalizer_p_1.pdf}}\label{fig:time_p1}}
    \subfloat[$p=16$]{{\includegraphics[width=.5\textwidth]{time_midaco_globalizer_p_16.pdf}}\label{fig:time_p16}}
    \caption{Averaged execution time of the parallel algorithms when solving problems from the GKLS 4d Hard class}
  \end{figure}
\end{frame}

%\begin{frame}
%  \frametitle{Results of applying the parallel algorithm}
%\end{frame}

\begin{frame}
  \frametitle{Conclusions}
    %Already done:
    \begin{itemize}
      \item 1
    \end{itemize}
    %Future work:
    %\begin{itemize}
    %  \item
    %\end{itemize}
\end{frame}
{
\unnumbered
\begin{frame}{{}}
  \frametitle{Q\&A}
  \begin{center}
    \Large{Contacts:}
\vspace{0.5cm}

>>>>

\begin{frame}
  \frametitle{Methods}
  \begin{itemize}
    \item[$\square$] \textbf{Deteriminstic}:
    \begin{itemize}
      \item Have complicated internal structure in multidimensional case;
      \item Usually store and use the whole history of trials accumulated during search;
      \item Require non-trivial parallelization schemes;
      \item Under some assumptions convergence to the global solution is guaranteed.
    \end{itemize}
    \item[$\square$] \textbf{Stochastic}:
    \begin{itemize}
      \item Have relative simple internal structure;
      \item Require constant amount of memory to store internal state of some random process or individuals of population;
      \item In most cases allows wide-scale parallelization by trials or by running many copies of the same method
      with different seeds in parallel.
      \item Convergence is guaranteed in probabilistic sense only.
    \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Methods}
  In this work considered the following solvers available in open-source:
  \begin{itemize}
    \item[$\square$] \textbf{Deteriminstic}:
    \begin{itemize}
      \item DIRECT, DIRECT$l$;
      \item AGS, AGS$l$.
    \end{itemize}
    \item[$\square$] \textbf{Stochastic}:
    \begin{itemize}
      \item Multi Level Single Linkage;
      \item StoGO;
      \item Differential Evolution;
      \item Controlled Random Search;
      \item Dual Simulated Annealing;
      \item Ant Colony Optimization.
    \end{itemize}
  \end{itemize}
\end{frame}

@@

\begin{frame}
  \frametitle{Univariate Algorithm of Global Search}
  Optimization method generates search sequence \(\{x_k\}\) and consists of the following steps:
  \begin{enumerate}
    \setlength{\itemindent}{.1in}
    \item[Step 1.] Sort the search information (one-dimensional points) in increasing order.
    \item[Step 2.] For each interval \((x_{i-1}, x_i)\) compute quantity \(R(i)\), called characteristic.
    \item[Step 3.] Choose \(p\) intervals \((x_{t_j-1}, x_{t_j})\) with the greatest characteristics and
    compute objective \(\varphi(y(x^{k+j}))\) in points chosen using the decision rule \(d\):
    \begin{displaymath}
      x^{k+1+j}=d(t)\in (x_{t_j-1}, x_{t_j}),\:j=\overline{1,p}
    \end{displaymath}
    \item[Step 4.] If \(x_{t_j}-x_{t_j-1}<\varepsilon\) for one of \(j=\overline{1,p}\), stop the method.
  \end{enumerate}
  \textit{\footnotesize	{Detailed description: Strongin R.G., Sergeyev Ya.D.: Global optimization with non-convex constraints. Sequential and parallel algorithms (2000), Chapter 7}}
\end{frame}

@@

\begin{frame}
  \frametitle{Globalizer parallel solver}
  \begin{figure}[ht]
    \includegraphics[width=0.95\textwidth]{evolvents_parallel.pdf}
  \end{figure}
\end{frame}

\begin{frame}
  \frametitle{MIDACO parallel solver}
  The MIDACO solver utilizes reverse communication architecture.
  Reverse communication means that the call of the objective function happens outside and
  independently of the MIDACO source code.
  Relying on this feature the following distributed parallelization scheme was implemented:
  \begin{enumerate}
    \item MIDACO runs at the master node and generates \(n \times p\) new trial points at each
  iteration.
    \item Each of \(n\) distributed computational nodes gets \(p\) points from the master and performs
  \(p\) parallel trials on local computing devices.
    \item Each of \(n\) distributed computational nodes sends \(p\) values of the objective function to
  the master.
  \end{enumerate}

  In this scheme all distributed communications were implemented using the MPI library.
  Parallelism within a single node is powered by the OpenMP standard.
\end{frame}

@@

\begin{frame}
  \frametitle{Results of the parallel algorithms}
  \begin{table}
    \centering
    \caption{Averaged numbers of iterations executed by the parallel algorithms for solving the test
  optimization problems}
    \label{tab:iterations}
    \begin{tabular}{cccccccc}
      \cline{3-8}\noalign{\smallskip}
      \multicolumn{2}{c}{  } & \textit{p} & \multicolumn{2}{c}{Globalizer} & &
  \multicolumn{2}{c}{MIDACO}   \\
      \noalign{\smallskip} \cline{4-5} \cline{7-8}  \noalign{\smallskip}
      \multicolumn{2}{c}{  } & & \textit{Simple} & \textit{Hard} & & \textit{Simple} &
  \textit{Hard}  \\
      \noalign{\smallskip}\hline
      I & \textbf{1 cluster node}
        & \textit{1} &   25270 (100)  & 55180 (99) & & 27645 (98) & 72068 (71)  \\
      &  & \textit{16} & 1765 (100)  & 3714 (100) & &  1640 (97) & 4304 (70) \\
      \hline \noalign{\smallskip}
  II  & \textbf{2 cluster nodes}  %\multirow{3}{*}{}
    & \textit{1} & 13056 (100) & 22938 (99) & & 10558 (89) & 27128 (73) \\
  &   & \textit{16} & 732 (100) & 1759 (100)  &  & 1130 (92) & 2254 (73) \\
      \hline \noalign{\smallskip}
  III & \textbf{4 cluster nodes} %\multirow{3}{*}{}
    & \textit{1}  & 5016 (100) & 12703 (100) & & 5777 (94) & 15980 (75) \\
  & & \textit{16} & 367 (100) & 776 (100) & & 529 (88) &  1264 (66) \\
      \hline \noalign{\smallskip}
  VI & \textbf{8 cluster nodes} %\multirow{3}{*}{}
    & \textit{1}  & 2103 (100) & 5063 (100) & & 2847 (97) & 9853 (83)\\
  & & \textit{16} & 145 (100)  & 310 (100)  & & 272 (83) & 774 (57)\\
      \hline \noalign{\smallskip}
  V & \textbf{12 cluster nodes} %\multirow{3}{*}{}
    & \textit{1}  & 1155 (100) & 2399 (100) & & 2233 (98) & 6022 (86) \\
  & & \textit{16} & 76 (100)  & 159 (100)  & & 168 (87) & 393 (53)\\
      \noalign{\smallskip}\hline
    \end{tabular}
  \end{table}
\end{frame}

\begin{frame}
  \frametitle{Results of the parallel algorithms}
  \begin{table}
    \centering
    \caption{Speedup of parallel computations executed by the parallel algorithms}
    \label{tab:speedup}
    \begin{tabular}{cccccccc}
      \cline{3-8}\noalign{\smallskip}
      \multicolumn{2}{c}{  } & \textit{p} & \multicolumn{2}{c}{Globalizer} & &
  \multicolumn{2}{c}{MIDACO}   \\
      \noalign{\smallskip} \cline{4-5} \cline{7-8}  \noalign{\smallskip}
      \multicolumn{2}{c}{  } & & \textit{Simple} & \textit{Hard} & & \textit{Simple} &
  \textit{Hard}  \\
      \noalign{\smallskip}\hline
  I  & \textbf{1 cluster node}  %\multirow{3}{*}{}
      & \textit{1}   & 25270 (27.3s) & 55180 (61.8s) & & 27645 (32.2s) & 72068 (132.5s)  \\
    &  & \textit{16} & 14.3(11.7) & 14.9(12.6)  & &  16.9(14.4) & 16.7(14.4) \\
    \hline \noalign{\smallskip}
  II  & \textbf{2 cluster nodes}  %\multirow{3}{*}{}
    & \textit{1}      &   1.9(1.9) & 2.4(2.2)  & & 2.6(1.7) & 2.7(2.3) \\
    &   & \textit{16} & 34.5(20.4) & 31.4(18.8) & & 24.5(18.8) & 32.0(29.1) \\
      \noalign{\smallskip}\hline	\noalign{\smallskip}
  III  & \textbf{4 cluster nodes}  %\multirow{3}{*}{}
  & \textit{1}      & 5.0(4.6) & 4.3(4.1) & & 4.8(3.9) & 4.5(4.4) \\
  &   & \textit{16} & 68.8(23.7) & 71.2(25.8) & & 52.3(32.9) & 57.0(50.2) \\
    \noalign{\smallskip}\hline	\noalign{\smallskip}
  VI & \textbf{8 cluster nodes} %\multirow{3}{*}{}
  & \textit{1}    & 12.0(8.7) & 10.9(8.3) & & 9.7(8.2)    & 7.3(9.0)    \\
  & & \textit{16} & 174.1(4.7) & 177.8(5.3) & & 101.6(51.6) & 93.1(84.1) \\
      \noalign{\smallskip}\hline
      V & \textbf{12 cluster nodes} %\multirow{3}{*}{}
      & \textit{1}    & 21.9(11.4)  & 23.0(12.8)  & & 12.4(11.1)  & 12.0(14.8)  \\
      & & \textit{16} & 333.5(2.7) & 347.9(3.0) & & 164.5(82.0) & 183.2(129.7) \\
          \noalign{\smallskip}\hline
    \end{tabular}
  \end{table}
\end{frame}

\begin{frame}
  \frametitle{Results of the parallel algorithms}
  \begin{figure}[ht]
    \centering
    \subfloat[$p=1$]{{\includegraphics[width=.5\textwidth]{speedup_midaco_globalizer_p_1.pdf}}\label{fig:speedup_p1}}
    \subfloat[$p=16$]{{\includegraphics[width=.5\textwidth]{speedup_midaco_globalizer_p_16.pdf}}\label{fig:speedup_p16}}
    \caption{Speedup demonstrated by the parallel algorithms when solving problems from the GKLS 4d Hard class}
  \end{figure}
\end{frame}

\begin{frame}
  \frametitle{Results of the parallel algorithms}
  \begin{figure}[ht]
    \centering
    \subfloat[$p=1$]{{\includegraphics[width=.5\textwidth]{time_midaco_globalizer_p_1.pdf}}\label{fig:time_p1}}
    \subfloat[$p=16$]{{\includegraphics[width=.5\textwidth]{time_midaco_globalizer_p_16.pdf}}\label{fig:time_p16}}
    \caption{Averaged execution time of the parallel algorithms when solving problems from the GKLS 4d Hard class}
  \end{figure}
\end{frame}

\begin{frame}
  \frametitle{Conclusions}
    %Already done:
    \begin{itemize}
      \item AGS\(l\) method implemented in the Globalizer system has demonstrated the convergence
    speed and reliability at the level of DIRECT and exceeds many other algorithms, the open-source
    implementations of which are available;
    \item the stochastic optimization methods inferior to the deterministic ones in the convergence
  speed and in reliability. It is manifested especially strongly on more complex multiextremal
  problems;
    \item the parallel version of the Globalizer system demonstrates good speedup
  when running on several nodes, on each of which a single objective function value per iterations is
  computed. When solving the problems with fast computable objective functions and using multiple
  threads on the nodes, the speedup for the Globalizer system degrade with increasing
  number of nodes;
    \item the MIDACO system is the most suitable for simple global optimization problems with the
  fast-computable objective functions. In this case, MIDACO is reliable enough and provides a linear
  speedup with increasing number of nodes and threads executed on these ones in parallel.
    \end{itemize}
    %Future work:
    %\begin{itemize}
    %  \item
    %\end{itemize}
\end{frame}
{
\unnumbered
\begin{frame}{{}}
  \frametitle{Q\&A}
  \begin{center}
    \Large{Contacts:}
\vspace{0.5cm}
